---
title: "BIS15L_AI_class"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
Today, we're exploring how AI can aid in writing, comprehending and troubleshooting our R code. AI is a very valuable tool, but it's important not to rely on it entirely. Building a strong R foundation is still important. These AI tools are not perfect and can be inaccurate solutions. In addition these tools can't read our minds, so understanding R helps us generate better prompts to get better responses. In short, AI should be used as a tool, not a crutch. Additionally, do not forget about other useful resources like Stack Overflow/Exchange, online tutorials, and GitHub issues pages for learning and troubleshooting. Let's combine AI with traditional methods to master R effectively.

This overview isn't exhaustive of all possible AI tools for R coding, but I believe it is a good start and covers everything you need.


**Make sure to enable Github Copilot in RStudio.**
Go to Tools -> Global Options -> Copilot -> Enable Git

## Learning Goals: 
*At the end of this exercise, you will be able to:*  
1. Explain what AI is and what AI tools can be used for R.
2. Enable Github Copilot in RStudio.
3. Use Copilot/ChatGPT to write code in RStudio.
4. Use Copilot/ChatGPT for troubleshooting code in RStudio.


## Exploration Questions:
Lets get some get a feel for what copilot (or ChatGPT) does, use Github Copilot to autocomplete the questions below. You can alternatively paste the questions in ChatGPT or Microsoft Copilot.
1. What is AI?
AI refers to the development of computer systems capable of performing tasks that typically require human intelligence, such as learning from data, reasoning, and solving problems.
2. What is ChatGPT?
ChatGPT is an AI conversational agent developed by OpenAI, utilizing the GPT architecture to engage in natural language conversations, answer questions, and perform various language tasks based on vast internet-trained text data.
3. What is Github Copilot?
GitHub Copilot is an AI-powered code completion tool developed by GitHub in collaboration with OpenAI, designed to assist developers by providing contextually relevant code suggestions and completions within their Integrated Development Environment (IDE) based on the code they are writing, comments, and documentation.

4. What can I use these tools for?
You can use these tools for coding help, learning, writing assistance, problem-solving, boosting productivity, and collaborating with others.
## Load the libraries  
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(dplyr)
library(shiny)
library(shinydashboard)
```

## Load the Data: 
```{r}
#load the Netflix Data in your Data folder, save it under the object name netflix
netflix <- read_csv("data/netflix_titles.csv") %>% clean_names()

```

*Did copilot help you write the code above? Could anything about the prompt be improved to help Copilot understand what you want to do more accurately?*
No.

## Lets Practice:
Explore the elements in your data: include a comment within your code chunk to prompt Copilot to use the function of your choice
```{r}
glimpse(netflix)
```
*Did copilot use your preferred method to get the summary of the data? If not, how could you improve the prompt to get the desired output?*


Sort out the movies from the netflix into an new object called netflix_movies
```{r} 
netflix_movies <- netflix %>%
  filter(type == "Movie")
```


Which director has the most movies on netflix?
```{r}
count(netflix_movies, director, sort = T)

```
*Do you see anything wrong with the table above, look at row 1. Write a prompt below to fix this.*

[insert your prompt here]
```{r}
filter(count(netflix_movies, director, sort = T), !is.na(director))
```



Which director had the most movies by release year and remove the NA's?
```{r}


```
*See anything else interesting about this table, look at the first director on the table. Could this be a problem and how can we edit the dataset to fix this? Try writing a prompt.*


separate the directors into their own rows if theres a comma presemt
```{r}
netflix_movies <- separate_rows(netflix_movies, director, sep = ",")
```

```{r}
netflix_movies <- netflix_movies %>%
  filter(!is.na(director))
```


Make a bar plot of the top directors with the most movies on netflix
```{r}
director_counts <- count(netflix_movies, director, sort = TRUE)
```

```{r}
ggplot(head(director_counts, 10), aes(x = reorder(director, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Top Directors with the Most Movies on Netflix",
       x = "Director",
       y = "Number of Movies") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
*Do you see any issues with this plot? How could we improve the prompt to get a better boxplot? Try writing a prompt.*

[Insert your prompt here]
```{r}

```

*How about for just 2020*



[Insert your prompt here]
```{r}
netflix_movies_2020 <- filter(netflix_movies, release_year == 2020)

# Count the occurrences of each director for the year 2020
director_counts_2020 <- count(netflix_movies_2020, director, sort = TRUE)

# Plot the top directors with the most movies on Netflix for 2020
ggplot(head(director_counts_2020, 10), aes(x = reorder(director, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Top Directors with the Most Movies on Netflix (2020)",
       x = "Director",
       y = "Number of Movies") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


*Take a few minutes to make your own plot(s) using prompts, edit prompts until you get a plot that you like.*






## Lets make the top 10 directors barplot into a Shiny App: use copilot or Chatgpt
```{r}
ui <- fluidPage(
  titlePanel("Top Directors with the Most Movies on Netflix"),
  sidebarLayout(
    sidebarPanel(
      selectInput("year", "Select Year", choices = unique(netflix_movies$release_year))
    ),
    mainPanel(
      plotOutput("barplot")
    )
  )
)

# Server
server <- function(input, output) {
  
  # Filter data based on selected year
  selected_data <- reactive({
    filter(netflix_movies, release_year == input$year)
  })
  
  # Count the occurrences of each director based on selected year
  director_counts <- reactive({
    count(selected_data(), director, sort = TRUE)
  })
  
  # Plot the top directors with the most movies based on selected year
  output$barplot <- renderPlot({
    ggplot(head(director_counts(), 10), aes(x = reorder(director, n), y = n)) +
      geom_bar(stat = "identity", fill = "purple", alpha = 0.8) +
      labs(title = paste("Top Directors with the Most Movies on Netflix (", input$year, ")"),
           x = "Director",
           y = "Number of Movies") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)
```


*Now use ChatGPT, Copilot or previous notes to fix the code*





## That's it! Let's take a break and then move on to working on group projects.
