---
title: "lab_6_warmup"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the Library

```{r}
library("tidyverse")
```


```{r}
library("janitor")
```


## Load the bison data
```{r}
bison <- read_csv("data/bison.csv")
```
## What are the dimensions and structure of the data?
```{r}
dim(bison)
```

```{r}
glimpse(bison)
```

## We are only interested in code, sex, weight, year of birth
```{r}
names(bison)
```

## Restrict tthe data to these variables and store as a new object
```{r}
bison_subset <- select(bison, "data_code", "animal_sex", "animal_weight", "animal_yob")
bison_subset
```
```{r}
#using pipes
#bison_new <- bison %>%
#  select(animal_code, animal_sex. animal_weight, animal_yob)
```

## Pull out the animals born between 1980-1990
```{r}
bison_year <- filter(bison_subset, animal_yob <= 1990 & animal_yob >= 1980)
bison_year
#could also use between
```
```{r}
#bison_new %>%
  #filter(animal_yob>= 1980 & animal_yob<= 1990)
```


## How many male and female bison are represented between 1980-1990?
There are 21 male bison represented and 414 female bison represented. 
```{r}
male_bison <- filter(bison_year, animal_sex == "M")
male_bison
```
```{r}
female_bison <- filter(bison_year, animal_sex == "F")
female_bison
```
```{r}
## class way
#males <- bison_new %>%
 # filter(animal_yob>= 1980 & animal_yob<= 1990) %>%
  #filter(animal_sex == "M")
#males

#table(males$animal_sex)
```

```{r}
## class way
#females <- bison_new %>%
  #filter(animal_yob>= 1980 & animal_yob<= 1990) %>%
  #filter(animal_sex == "M")
#females

#table(females$animal_sex)
```

## Between 1980-1990, were males or females larger on average?

On average it appears that the male bison were larger. 
```{r}
mean(male_bison$animal_weight)
```

```{r}
mean(female_bison$animal_weight)
```

